Summer 2025 ToDo

### 2025-10-15 Relations/Paper/Well-foundedness.tex
getting close to finishing section 5

\begin{definition} \label{def:WFproperties} \hfill 
  \begin{enumerate}
    \item $R$ is \emph{accessibly well-founded} ($\WFacc$) if every element is accessible:
      \[
        \forall x \in A.\; x \text{ is accessible}
      \]
    \item $R$ is \emph{inductively well-founded} ($\WFind$) if every inductive predicate is universally true:
      \[
        \forall P \subseteq A.\;\text{$P$ is inductive} \Rightarrow \forall x \in A.\; x \in P
      \]
    \item $R$ is \emph{coreductively well-founded} ($\WFcor$) if every coreductive predicate is universally true:
      \[
        \forall P \subseteq A.\;\text{$P$ is coreductive} \Rightarrow \forall x \in A.\; x \in P
      \]
    \item  $R$ is \emph{well-founded minimality-wise} ($\WFmin$) if every nonempty subset contains a minimal element:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow
        \exists x \in A.\; \text{$x$ is $P$-minimal}
      \]
    \item  $R$ is \emph{well-founded minimality-wise for $\lnot \lnot$-closed predicates} ($\WFminDNE$) if every nonempty $\lnot \lnot$-closed subset contains a minimal element:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow \ol{\ol P} \subseteq P \Rightarrow
        \exists x \in A.\; \text{$x$ is $P$-minimal}
      \]
   \item $R$ is \emph{sequentially well-founded} ($\WFseq$) if every sequence contains an index at which it is not decreasing:
      \[
        \forall s : \nat \to A\ \exists k : \nat.\, \lnot \, R \,(s\,(k+1))\,(s\,k)
      \]
  \end{enumerate}
\end{definition}
We have formalized the above definitions of well-foundedness in \texttt{WFDefinitions.agda}.

@@

% As the figure illustrates, the concept of well-foundness is very rich from the vantage point of constructive logic.
% Our analysis resulted in the following principles, 
%
% The additional classical hypotheses needed to recover some of the implications on the diagram 
%
% The principles appearing in the labels of Figure \ref{fig:WF} are valid for many rewriting systems,
% and are classically true.  
%
% The weaker definitions help us to explore how the standard well-founded definitions interrelate. This is made clear in the following diagram.
%
As can be seen from the figure, taking the $\lnot\lnot$-closures of 
well-foundedness properties cuts down some, but not all, of the complexity.
For example, the two minimality principles
described above are equivalent with this formulation.
% The implication from the inductive to sequential definition no longer requires
% decidability of $R$.
% On the other hand, the reverse
At the same time, the weak version of the sequential definition is unable to imply 
the (weak version of) accessible one without an additional hypothesis $\accCor$, 
which may be described as the constructive contrapositive of the definition of accessibility.
In the context of rewriting systems,
this condition asserts the existence of a function which maps any
non-strongly normalizing element to a one-step reduct with the same property.
Assuming $\accCor$, all notions of well-foundedness
are constructively equivalent up to $\lnot\lnot$.

As the Figure \ref{tab:cprop} illustrates, the network of relationships between the
constructive notions of well-foundedness is quite rich. 
The main observations can now be read off from the figure.
\begin{enumerate}
  \item The implication from $\isWFacc$ to $\isWFseq$ requires the underlying relation to be decidable. 
    This assumption is no longer necessary on the $\lnot\lnot$-translated side.
  \item WFmin is a very strong assumption and implies WFseq unconditionally.
  The more reasonable WFminDNE requires the image of every sequence to be $\lnot\lnot$-closed 
  to complete the implication.  This condition can be considered as a special case of Markov's Principle
  (assuming decidability of equality on $A$).
  \item On the $\lnot\lnot$-side, WFmin- and WFminDNE- are equivalent,
  and imply WFacc- via WFseq-, requiring {$\accCor$} assumption in the final inference.
  \item At the same time, if the relation is finitely branching ($\FB$), then WFminDNE-
  implies WFacc- directly.  It should be noted that, $\FB$ together with (weak) decidability of minimality
  implies $\accCor$.  \apcomment{Fix this: This is proved in \texttt{FinitelyBranching.agda}.}.
  % \apcomment{The previous observation suggests the question of whether isWFseq- might
  % imply isWFminDNE- via decidability of minimal elements.\\
  % However, in trying to prove this implication, the hypothesis asserting non-existence of
  %  a minimal element of some given non-empty $\lnot\lnot$-closed predicate
  %  automatically entails that any particular element of interest is not normal.\\
  %  Unless it's not satisfying $P$? \emph{We need to decide whether it is P-minimal as well.}}
  \item Assuming $\accCor$, $\isWFacc$ ($\isWFaccm$) can also be derived from a weaker condition, $\WFcor$ ($\WFcorm$).
    \apcomment{Add the ``is''!}
    (Just like $\isWFminDNE$ is a natural weakening of $\isWFmin$ that is sufficient to recover $\isWFacc$ in the presence of {$\accDNE$}.)
  \apcomment{True?  It is implied by $\isWFseqm$, but does not appear to imply anything on its own.}

  % \apcomment{(This seems easy: The converse implication isWFminCor to isWFseq- requires the sequence in question to
  % have decidable image, leading to a further weakening of isWFseq-.)}
\item \apcomment{Fix with macros}. Since isWFminDNE- implies isWFseq-, the former also implies isWFminCor-.
  But the converse fails: If a subset/predicate $P$ is $\lnot\lnot$-closed,
  we cannot conclude it is coreductive.

  \item In general, the coreductive properties speak to the ability to choose
  a successor (strong existence).  This ability is central to being able to
  define sequences.  Therefore, sequential definitions generally require
  some coreductive situation to be applicable.

@@

In \cite{Terese}, well-foundedness is defined as in the classical mathematics textbooks,
as the absence of an infinite $R$-decreasing sequence.
(Again, well-foundedness of $R$ corresponds to strong normalization of its converse.)
Translating this statement directly into type theory yields the formula

>>>>

\begin{definition} \label{def:WFproperties} \hfill 
  \begin{enumerate}
    \item $R$ is \emph{accessibly well-founded} ($\WFacc$) if every element is accessible:
      \[
        \forall x \in A.\; x \text{ is accessible}
      \]
    \item \label{def:WFind} $R$ is \emph{inductively well-founded} ($\WFind$) if every inductive predicate is universally true:
      \[
        \forall P \subseteq A.\;\text{$P$ is inductive} \Rightarrow \forall x \in A.\; x \in P
      \]
    \item $R$ is \emph{coreductively well-founded} ($\WFcor$) if every coreductive predicate is universally true:
      \[
        \forall P \subseteq A.\;\text{$P$ is coreductive} \Rightarrow \forall x \in A.\; x \in P
      \]
    \item  $R$ is \emph{well-founded minimality-wise} ($\WFmin$) if every nonempty subset contains a minimal element:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow
        \exists x \in A.\; \text{$x$ is $P$-minimal}
      \]
    \item  $R$ is \emph{well-founded minimality-wise for $\lnot \lnot$-closed predicates} ($\WFminDNE$) if every nonempty $\lnot \lnot$-closed subset contains a minimal element:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow \ol{\ol P} \subseteq P \Rightarrow
        \exists x \in A.\; \text{$x$ is $P$-minimal}
      \]
   \item $R$ is \emph{sequentially well-founded} ($\WFseq$) if every sequence contains an index at which it is not decreasing:
      \[
        \forall s : \nat \to A\ \exists k : \nat.\, \lnot \, R \,(s\,(k+1))\,(s\,k)
      \]
  \end{enumerate}
\end{definition}
We have formalized the above definitions of well-foundedness in \texttt{WFDefinitions.agda}.

@@

\apcomment{Make another version of the diagram, swapping WFmin and WFacc rows.}

% The principles appearing in the labels of Figure \ref{fig:WF} are valid for many rewriting systems,
% and are classically true.  
Figure \ref{tab:cprop} illustrates that the network of logical relationships between
constructive notions of well-foundedness is quite rich indeed. 
Taking the $\lnot\lnot$-closures of 
well-foundedness properties cuts down some, but not all, of the complexity.
For example, the two minimality principles become equivalent.
% The implication from the inductive to sequential definition no longer requires
% decidability of $R$.
% On the other hand, the reverse
At the same time, the weak version of the sequential definition is unable to imply 
the (weak) accessible one without an additional hypothesis $\accCor$, 
which may be described as the constructive contrapositive of the definition of accessibility.
In the context of rewriting systems,
this condition asserts the existence of a function which maps any
element that is \emph{not} strongly normalizing to a one-step reduct of it having the same property.
Assuming $\accCor$, all notions of well-foundedness
are constructively equivalent up to $\lnot\lnot$.

Other observations that can be gleaned from the figure include the following.
\apcomment{The macros have to fixed throughout.}
\begin{enumerate}
  \item The implication from $\isWFacc$ to $\isWFseq$ requires the underlying relation to be decidable. 
    This assumption is no longer necessary on the $\lnot\lnot$-translated side.
  \item WFmin is a very strong assumption and implies WFseq unconditionally.
  The more reasonable WFminDNE requires the image of every sequence to be $\lnot\lnot$-closed 
  to complete the implication.  This condition can be considered as a special case of Markov's Principle
  (if one assumes that equality on $A$ is decidable).
  \item On the $\lnot\lnot$-side, WFmin- and WFminDNE- are equivalent, are directly implied by $\isWFaccm$, 
    and directly imply $\isWFseqm$  \apcomment{Fix macros}.
  \item All definitions imply $\WFcorm$.  This weakest definition implies $\isWFaccm$ assuming the 
    ``constructive contrapositive of accessibility'', $\accCor$.
    %While this condition is implied by $\isWFseqm$, it does not appear to imply anything on its own.
  \item Assuming $\accCor$, $\isWFacc$ can also be derived from $\WFcor$. % , the stronger version of $\WFcorm$.
    \apcomment{Add the ``is''!}

    (This is analogous to how $\isWFminDNE$ is a natural weakening of 
    $\isWFmin$ that is sufficient to recover $\isWFacc$ in the presence of {$\accDNE$}.)

  \item At the same time, if the relation is finitely branching ($\FB$), then WFminDNE-
    implies WFacc- directly.

    (It may be noted that $\FB$ implies $\accCor$ whenever the relation $R$ is decidable
    and $R$-accessibility is \emph{weakly} decidable;  this is proved in \texttt{Coreductive.agda}.)
  % \apcomment{The previous observation suggests the question of whether isWFseq- might
  % imply isWFminDNE- via decidability of minimal elements.\\
  % However, in trying to prove this implication, the hypothesis asserting non-existence of
  %  a minimal element of some given non-empty $\lnot\lnot$-closed predicate
  %  automatically entails that any particular element of interest is not normal.\\
  %  Unless it's not satisfying $P$? \emph{We need to decide whether it is P-minimal as well.}}
  % \apcomment{(This seems easy: The converse implication isWFminCor to isWFseq- requires the sequence in question to
  % have decidable image, leading to a further weakening of isWFseq-.)}

% \item \apcomment{Fix with macros}. Since isWFminDNE- implies isWFseq-, the former also implies isWFminCor-.
%   But the converse fails: If a subset/predicate $P$ is $\lnot\lnot$-closed,
%   we cannot, in general, conclude it is coreductive.
  \item \apcomment{Delete? In general, the coreductive properties speak to the ability to choose
  a successor (strong existence).  This ability is central to being able to
  define sequences.  Therefore, sequential definitions generally require
some coreductive situation to be applicable.}

@@

We have also considered the statement that the complement of every coreductive predicate
  has a least element.  This statement and its $\lnot\lnot$-version are both equivalent to $\WFcorm$
  (see \texttt{CoreductiveImplications.agda}), so we have not included them in the figure.

We also did not include the inductive definition, Definition \ref{def:WFproperties}.\ref{def:WFind}.
  Logically, it is trivially equivalent to the accessible one, but its Agda formulation has a defect: 
  the inductive well-foundedness property lands in the larger universe $\mathtt{Set}_1$, due to 
  quantification over all predicates on $A$.  (This defect is also present in $\isWFmin$, $\isWFminDNE$, and \texttt{isWFcor}.)

In \cite{Terese}, well-foundedness is defined as in the classical mathematics textbooks,
as the absence of an infinite $R$-decreasing sequence.
(Again, well-foundedness of $R$ corresponds to strong normalization of its converse.)
\apcomment{Add a discussion about isWFseq- vs. isWFseq$\lnot\lnot$.}

We consider $\isWFacc$ the canonical constructive definition of a well-founded relation.

\apcomment{Todo: 1. Filter through the rest, and see if there's anything we may still want to keep.\\
  2. Include a subsection, or a section, with examples.  Provide examples where some decidability 
properties are provable, and others are not.}

\subsection{Old. Delete?}

Translating this statement directly into type theory yields the formula
